---
title: tidyverse
author: Michael Levy, Prepared for the Davis R-Users' Group
date: October 13, 2016
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```


## What is the tidyverse?

~~Hadleyverse~~

The tidyverse is a suite of R tools that follow a tidy philosophy:

### Tidy data

Put data in data frames  

- Each type of observation gets a data frame
- Each variable gets a column
- Each observation gets a row

### Tidy APIs

Functions should be consistent and easily (human) readable

- Take one step at a time
- Connect simple steps with the pipe
- Referential transparency


### Okay but really, what is it? 

Suite of ~20 packages that provide consistent, user-friendly, smart-default tools to do most of what most people do in R.

- Core packages: ggplot2, dplyr, tidyr, readr, purrr, tibble
- Specialized data manipulation: hms, stringr, lubridate, forcats
- Data import: DBI, haven, httr, jsonlite, readxl, rvest, xml2
- Modeling: modelr, broom

`install.packages(tidyverse)` installs all of the above packages.

`library(tidyverse)` attaches only the core packages.


## Why tidyverse?

- Consistency  
    - e.g. All `stringr` functions take string first  
    - e.g. Many functions take data.frame first -> piping
        - Faster to write
        - Easier to read
    - Tidy data: Imposes good practices
    - Type specification
- You probably use some of it already. Synergize.
- Implements simple solutions to common problems (e.g. `purrr`!)
- Smarter defaults 
    - e.g. `utils::write.csv(row.names = FALSE)` = `readr::write_csv()` 
- Runs fast (thanks to `Rcpp`)
- Interfaces well with other tools (e.g. Spark with `dplyr` via `sparklyr`)

## `tibble`

> A modern reimagining of data frames.

```{r Attach core packages}
library(tidyverse)
```

```{r class tbl}
tdf = tibble(x = 1:1e4, y = rnorm(1e4))  # == data_frame(x = 1:1e4, y = rnorm(1e4))
class(tdf)
```


Tibbles print politely. 

```{r print tbl}
tdf
```


- Can customize print methods with `print(tdf, n = rows, width = cols)`

- Set default with `options(tibble.print_max = rows, tibble.width = cols)`

Tibbles have some convenient and consistent defaults that are different from base R data.frames.

#### strings as factors

```{r strings as factors}
dfs = list(
  df = data.frame(abc = letters[1:3], xyz = letters[24:26]),
  tbl = data_frame(abc = letters[1:3], xyz = letters[24:26])
)
sapply(dfs, function(d) class(d$abc))
```


#### partial matching of names

```{r partial matching}
sapply(dfs, function(d) d$a)
```

#### type consistency

```{r single bracket excision}
sapply(dfs, function(d) class(d[, "abc"]))
```

Note that tidyverse import functions (e.g. `readr::read_csv`) default to tibbles and that *this can break existing code*.

#### List-columns!

```{r, list columns}
tibble(x = lapply(1:5, function(x) x^(1:x)))
```


## The pipe `%>%`

Sends the output of the LHS function to the first argument of the RHS function.

```{r pipe}
sum(1:8) %>%
  sqrt()
```


## `dplyr`

Common data(frame) manipulation taks. 

Four core "verbs": filter, select, arrange, group_by + summarize, plus many more convenience functions. 


```{r load movies}
library(ggplot2movies)
str(movies)
```

```{r filter}
filter(movies, length > 360)
```

```{r select}
filter(movies, length > 360) %>%
  select(title, rating, votes)
```

```{r arrange}
filter(movies, Animation == 1, votes > 1000) %>%
  select(title, rating) %>%
  arrange(desc(rating))
```

`summarize` makes `aggregate` and `tapply` functionality easier, and the output is always a data frame.

```{r summarize}
filter(movies, mpaa != "") %>%
  group_by(year, mpaa) %>%
  summarize(avg_budget = mean(budget, na.rm = TRUE),
            avg_rating = mean(rating, na.rm = TRUE)) %>%
  arrange(desc(year), mpaa)
```


`count` for frequency tables. Note the consistent API and easy readability vs. `table`.

```{r count}
filter(movies, mpaa != "") %>%
  count(year, mpaa, Animation, sort = TRUE)
```


```{r table}
basetab = with(movies[movies$mpaa != "", ], table(year, mpaa, Animation))
basetab[1:5, , ]
```



## `tidyr`

Latest generation of `reshape`. `gather` to make wide table long, `spread` to make long tables wide.

```{r}
who
```

```{r}
who %>%
  gather(group, cases, -country, -iso2, -iso3, -year)
```


## `ggplot2`

If you don't already know and love it, check out [one of](https://d-rug.github.io/blog/2012/ggplot-introduction) [our](https://d-rug.github.io/blog/2013/xtsmarkdown) [previous](https://d-rug.github.io/blog/2013/formatting-plots-for-pubs) [talks](https://d-rug.github.io/blog/2015/ggplot-tutorial-johnston) on ggplot or any of the excellent resources on the internet. 

Note that the pipe and consistent API make it easy to combine functions from different packages, and the whole thing is quite readable.

```{r dplyr-tidyr-ggplot, fig.width = 4, fig.asp = 1}
who %>%
  select(-iso2, -iso3) %>%
  gather(group, cases, -country, -year) %>%
  count(country, year, wt = cases) %>%
  ggplot(aes(x = year, y = n, group = country)) +
  geom_line(size = .2) 
```


## `readr`

For reading flat files. Faster than base with smarter defaults.

```{r}
bigdf = data_frame(int = 1:1e6, 
                   squares = int^2, 
                   letters = sample(letters, 1e6, replace = TRUE))
```

```{r base write}
system.time(
  write.csv(bigdf, "base-write.csv")
)
```

```{r readr write}
system.time(
  write_csv(bigdf, "readr-write.csv")
)
```

```{r base read}
read.csv("base-write.csv", nrows = 3)
```

```{r readr read}
read_csv("readr-write.csv", n_max = 3)
```



---

For stringr...

```{r}

library(stringr)
who %>%
    select(-iso2, -iso3) %>%
    gather(group, cases, -country, -year) %>%
    mutate(group = str_replace(group, "(new_)|(new)", ""),
           method = str_extract(group, "[a-z]+"),
           gender = str_sub(str_extract(group, "_[a-z]"), 2, 2),
           age_group = str_extract(group, "[0-9]+")) %>%
    group_by(country, year, method) %>%
    summarize(total_cases = sum(cases, na.rm = TRUE)) %>%
    ggplot(aes(x = year, y = total_cases, group = country)) +
    geom_line() +
    facet_wrap(~ method)
```
